\MOC{} Server Pages:

### 2007-04-16 paper/paper.tex
svn-r20: Added staged dir browser in PHP, and its performance measurements.
Improved figures and listings in paper.

darcs-hash:20070416222632-e5a07-32d60fb7a0f5085fabbba0a98521b9a1b3d1c01d.gz
svn-r21: Bug fix in translator, for when comments found in let block.  Changes to paper.
darcs-hash:20070416222703-e5a07-2caa368989cb0178a7e31f84c708dbd1d028758e.gz

\documentclass[preprint]{acm_proc_article-sp}
\usepackage{ucs,textcomp,url,graphicx,color}
\usepackage[leftno,lineno5]{lgrind}
\usepackage[bf]{caption2}
\usepackage[numbers]{natbib}
\usepackage[utf8]{inputenc}
\ifx\pdfoutput\undefined\else
  \usepackage[
    pdfpagemode=None,
    pdftitle={MetaOCaml Server Pages},
    pdfsubject={Web Publishing as Staged Computation},
    pdfauthor={Christopher League}
  ]{hyperref}
\fi
\input{fonts}
\def\MOC{MetaOCaml}
\hyphenation{Meta-OCaml data-base Slash-dot}
\def\quote{\list{}{\leftmargin1.4em\rightmargin1em}\item[]}
\newcommand{\myfig}[3]{%
  \begin{figure}[tbp]%
    #3%
    \caption{#2}%
    \label{fig:#1}%
  \end{figure}}
\newcommand{\screenshot}[2]{%
  \myfig{#1}{#2}
  {\centering%
    \includegraphics[width=20pc,keepaspectratio]{#1}}}
\newcommand{\gnuplot}[2]{%
  \myfig{#1}{#2}{\input{#1}}}
\newcommand{\listing}[2]{\lagrind[tbp]{#1}{#2}{lst:#1}}

@@

In their infancy (ca.~1993), web sites were little more than
clickable FTP sites with in-line images.  They enabled
distribution of static documents, but did little to enhance
collaboration or create a sense of community.  In contrast,
today's users expect most web sites to offer sophisticated
services and to permit endless personalization.  In 2004,
even a strictly personal web site is expected to incorporate
user-contributed content such as comments and ratings.
\newpage

@@

Declaration blocks have been lifted to the top; any side effects
contained there are executed when the page is published.  The
constructed code begins on line~\ref{line:page} of
listing~\ref{lst:trans.ml}.  It is a function of two arguments:
\cd{req} encapsulates the HTTP request, including the headers
and query arguments; \cd{puts} is a function with type
\cd{string\to unit}; it transmits text across the network to the
visitor's browser.  \cd{arg} looks up an
argument in the query string of the request; its type is
\cd{string\to string option}.  These are all \emph{serve} (second)
stage values.  It makes no sense to access them in the
\emph{publish} (first) stage, and indeed the \MOC{} type system
prevents this.

@@

Now the code can be instantiated with early parameters in countless
ways.  For example, requesting the URI \id{/longcount} could print a
count-down from 99, while \id{/shortcount} starts at 9 instead.  Both
are generated from the same server page.  Mapping from URIs to
instantiated \MOC{} code is, for now, an implementation detail
(see \S\ref{sec:impl}).

@@

To explore the expressiveness of this design, we now look at a
series of web services, organized to comprise a small web
site.
%\footnote{I hope to make the demo available on the web in time
%  for the workshop; this will require convincing the I.T. folks to
%  open up a non-standard port through the firewall.  For now, it may
%  be downloaded and run locally.} %
To give the services a similar look, we developed a site-wide style
sheet, and an OCaml function to generate a navigation bar from a
hierarchical list of page titles and links.  The following bindings are
available site-wide:

@@

Judging from its prevalence in the literature, we are certain that
millions of grateful users would subscribe to an online service
capable of computing the \emph{power} function!  The screen shot in
figure~\ref{fig:power127} illustrates how it works.  The navigation
bar shows the exponents for which code has been pre-generated.  It is
not possible, with the current implementation, for the user to request
other exponents (see \S\ref{sec:impl} for an explanation).  After
selecting the exponent, the user types the base into the form and
presses return.  The result is computed using OCaml's
arbitrary-precision \cd{Num} module.  The implementation of this
service appears in listing~\ref{lst:power.meta}.

@@

The first function wraps a string into a fixed width block.  The power
function itself begins on line~\ref{line:power-fun}.  No output should
occur until the preamble is generated on
line~\ref{line:power-preamble}.  The user's input is translated to a
number relatively late (line~\ref{line:power-num}); if there is a
problem here, the navigation bar and form will have already been
output before the exception message.

@@

Many web servers can be configured to permit clients to browse
directories.  The web server generates, on the fly, an HTML page
containing the names and attributes of (and hyperlinks to) all
the files in the directory.  In Apache, the
\texttt{mod\_autoindex} module provides this feature.

ViewCVS is a more complex example of the same idea.  It is a
Python program that sits on the web server and allows remote
users to browse a CVS repository with a standard web browser.

This kind of service can be fairly intensive; a single HTTP
request is likely to generate dozens of system calls and disk
accesses.  If the directory is viewed more often that it is
changed, then it makes sense to cache or pre-generate the pages.
It's simple enough to write a script to generate directory pages
off-line, but what if we also want dynamic user-controlled
behavior, such as sorting and filtering?

Our contribution to the genre not only lists files in a given
directory, but displays their MD5 checksums, renders their sizes
in human-readable form (140k or 1.2M), and colorizes their names
based on their extension or kind.  At serve time, the user may
specify a regular expression for filtering, and one of 5
criteria for sorting.  The screen shot in
figure~\ref{fig:server-dir} shows the result of browsing the
source directory of the server.

\screenshot{server-dir}{Browsing the server source code directory.}

@@

\mylst{dir.meta}{...}

@@

Some web servers can be configured to display their status in
response to certain URIs (such as \id{/server-status} on
Apache).  We programmed a few status services in \MOC.  The
screen shot in figure~\ref{fig:gc} displays garbage collection
statistics from the OCaml \id{GC} module.

\screenshot{gc}{Garbage collector statistics.}

The only complicated bit in the code is to format the numbers
using commas every three digits.  The descriptions are copied
directly from the \id{gc.mli} interface in OCaml.  The only
thing pre-generated here is the navigation bar (line 2).

@@

\gnuplot{power}{Staged vs.\@ unstaged power function.  The
  server page computes 2 to the given power
  (2\textsuperscript{127}, 2\textsuperscript{255}, etc.)}

@@

\gnuplot{browse}{Staged vs.~unstaged directory browsing.  *The
  Apache results refer to \texttt{mod\_autoindex}; it does
  \emph{not} compute the MD5 sum.}

@@

Figure~\ref{fig:browse} also includes the performance of
Apache's directory browser, \texttt{mod\_autoindex}.  It cannot
keep up with the staged browser, even though it doesn't compute
the MD5 sum or filter the files on demand.

@@

\newpage
\section{Future directions}
\label{sec:future}

@@

>>>>

\documentclass[preprint]{acm_proc_article-sp}
\usepackage{ucs,textcomp,url,xspace,graphicx,color}
\usepackage[leftno,lineno5]{lgrind}
\usepackage[bf]{caption2}
\usepackage[numbers]{natbib}
\usepackage[utf8]{inputenc}
\ifx\pdfoutput\undefined\else
  \usepackage[
    pdfpagemode=None,
    pdftitle={MetaOCaml Server Pages},
    pdfsubject={Web Publishing as Staged Computation},
    pdfauthor={Christopher League}
  ]{hyperref}
\fi
\input{fonts}
\def\MOC{MetaOCaml\xspace}
\hyphenation{Meta-OCaml data-base Slash-dot}
\def\quote{\list{}{\leftmargin1.4em\rightmargin1em}\item[]}
\newcommand{\myfig}[3]{%
  \begin{figure}[tbp]%
    #3%
    \caption{#2}%
    \label{fig:#1}%
  \end{figure}}
\newcommand{\screenshot}[2]{%
  \myfig{#1}{#2}
  {\centering%
    \includegraphics[width=20pc,keepaspectratio]{#1}}}
\newcommand{\gnuplot}[2]{%
  \myfig{#1}{#2}{\input{#1}}}
\newcommand{\listing}[2]{\lagrind[tbp]{#1}{#2}{lst:#1}}

@@

In their infancy (ca.~1993), web sites were little more than
clickable FTP sites with in-line images.  They enabled
distribution of static documents, but did little to enhance
collaboration or create a sense of community.  In contrast,
today's users expect most web sites to offer sophisticated
services and to permit endless personalization.  In 2004,
even a strictly personal web site is expected to incorporate
user-contributed content such as comments and ratings.

@@

Declaration blocks have been lifted to the top; any side effects
contained there are executed when the page is published.  The
constructed code begins on line~{???line:page} of
listing~\ref{lst:trans.ml}.  It is a function of two arguments:
\cd{req} encapsulates the HTTP request, including the headers
and query arguments; \cd{puts} is a function with type
\cd{string\to unit}; it transmits text across the network to the
visitor's browser.  \cd{arg} looks up an
argument in the query string of the request; its type is
\cd{string\to string option}.  These are all \emph{serve} (second)
stage values.  It makes no sense to access them in the
\emph{publish} (first) stage, and indeed the \MOC{} type system
prevents this.

@@

Now the code can be instantiated with early parameters in countless
ways.  For example, requesting the URI \id{/longcount} could print a
count-down from 99, while \id{/shortcount} starts at 9 instead.  Both
are generated from the same server page.  Mapping from URIs to
instantiated \MOC{} code is, for now, an implementation detail
(see section~\ref{sec:impl}).

@@

To explore the expressiveness of this design, we now look at a
series of web services, organized to comprise a small web
site.\footnote{A demo is available at
  \url{http://comsci.liu.edu/metaserv/}}
To give the services a similar look, we developed a site-wide style
sheet, and an OCaml function to generate a navigation bar from a
hierarchical list of page titles and links.  The following bindings are
available site-wide:

@@

Judging from its prevalence in the literature, we are certain that
millions of grateful users would subscribe to an online service
capable of computing the \emph{power} function!  The screen shot in
figure~\ref{fig:power127} illustrates how it works.  The navigation
bar shows the exponents for which code has been
pre-generated.\footnote{In the current implementation, it is not
  possible for the user to request other exponents once the server is
  running.  See section~\ref{sec:impl} for an explanation.}  After
selecting the exponent, the user types the base into the form and
presses return.  The result is computed using OCaml's
arbitrary-precision \cd{Num} module.  The implementation of this
service appears in listing~\ref{lst:power.meta}.

@@

The first function wraps a string into a fixed width block.  The power
function itself begins on line~{???line:power-fun}.  No output should
occur until the preamble is generated on
line~{???line:power-preamble}.  The user's input is translated to a
number relatively late (line~{???line:power-num}); if there is a
problem here, the navigation bar and form will have already been
output before the exception message.

@@

Many web servers can be configured to permit clients to browse
directories.  The web server generates, on the fly, an HTML page
containing the names and attributes of (and hyperlinks to) all
the files in the directory.  In Apache, the
\texttt{mod\_autoindex} module provides this feature.
%%
ViewCVS\footnote{\url{http://viewcvs.sourceforge.net/}} is a more
complex example of the same idea. It is a Python program that sits on
the web server and allows remote users to browse a CVS repository with
a standard web browser.

This kind of service can be fairly intense; one HTTP request is likely
to generate dozens of system calls and disk accesses.  If the
directory is viewed more often that it is changed, then it makes sense
to cache or pre-generate the pages.  Although it is simple enough to
write a script to generate directory pages off-line, what if we also
want dynamic user-controlled behavior, such as sorting and filtering?

We implemented
contribution to the genre not only lists files in a given
directory, but displays their MD5 checksums, renders their sizes in
human-readable form (`1.2M' rather than `1194822'), and colorizes their
names based on their extension or kind.  At serve time, the user may
specify a regular expression for filtering, and one of 5 criteria for
sorting.  The screen shot in figure~\ref{fig:server-dir} shows the
result of browsing the source directory of the server.

\screenshot{server-dir}{Browsing the server source code directory.
  File names are colored according to their \emph{kind,} which is
  computed from the extension and the file type.  See also
  listing~\ref{lst:dir.meta}.}

@@

\mylst{dir.meta}{This implements the directory browser shown in
  figure~\ref{fig:server-dir}.}

@@

Some web servers can be configured to display their status in response
to certain URIs (such as \id{/server-status} on Apache).  We
programmed a few status services in \MOC.  The screen shot in
figure~\ref{fig:gc} displays garbage collection statistics from the
OCaml \id{GC} module.  The only thing pre-generated here is the
navigation bar.

\screenshot{gc}{Garbage collector statistics.  The descriptions are
  copied directly from the \id{gc.mli} interface in OCaml.}

@@

\gnuplot{power}{Staged versus unstaged power function.  The
  server page computes 2 to the given power
  (2\textsuperscript{127}, 2\textsuperscript{255}, etc.)}

@@

\gnuplot{browse}{Staged versus unstaged directory browsing, in \MOC
  and PHP.}

@@

@@

\section{Future directions}
\label{sec:future}

@@

\section*{Acknowledgments}

Many thanks to the anonymous reviewers of the \MOC workshop, for
helpful recommendations on both the work and the presentation.
